Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is a merged and final version of two conference papers, \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}, with a unified and streamlined presentation.}

### 2020-07-14 journal paper/content/sec-theory.tex
Minor edits throughout, almost done.

We characterize the inherent quality of an MAB algorithm in terms of its \emph{Bayesian Instantaneous Regret} (henceforth, \BIR), a standard notion from machine learning:
\begin{align}\label{eq:theory-BIR-defn}
\BIR_i(n) := \E\sbr{ n\cdot \max_{a\in A} \mu_a - \rew_i(n)}.
\end{align}
We are primarily interested in how \BIR scales with $n$: typically, how fast \BIR decreases with $n$. We treat $K$, the number of arms, as a constant unless specified otherwise.
%Let $\BIR_i(n) = \BIR(n)$ for the algorithm of principal $i$.

Let us connect \BIR with the three classes of algorithms listed in Section~\ref{sec:intro-model}. \DynGreedy has at least a constant \BIR for many reasonable priors (where the constant can depend on $K$ and the prior, but not on $n$). \StaticGreedy typically has at least a constant \BIR, in the same sense. Exploration-separating algorithms can have
    $\BIR(n) \sim n^{-1/3}$
for all priors, whereas adaptive-exploration algorithms can have an even better rate
    $\BIR(n) \sim n^{-1/3}$
for all priors. Both rates are up to $\log(n)$ factors, and are the best possible \BIR rates that can be achieved for all priors. More details on this can be found in Appendix~\ref{app:examples}.

@@

\ascomment{Steven: do you buy this example? }

@@

\ascomment{Steven: OLD EXAMPLE. I revised it to clarify what it is supposed to mean. But my new example is better, isn't it?}

\ascomment{The conclusion can't be right: "greedy modification" can only improve the BIR at step $n_0$! Perhaps what we derive is that
$\BIR_2(n)>\BIR_1(n)$ starting from $n=n_0+1$? }

\begin{example}
Consider \alg[1] and \alg[2] as in Theorem~\ref{thm:random-greedy}.
Let $\BIRgr(n)$ be the \BIR of the ``greedy choice" after $n-1$ steps of \alg[1]. Assume, for some absolute constants $\beta,\gamma>0$ and $c>1-\gamma$, that
\begin{align}\label{eq:cor:random-greedy-ex-assumption}
 \tfrac{1}{c}\; \BIRgr(n) \geq \BIR_1(n)= \beta\cdot n^{-\gamma}
 \qquad \text{for all local steps $n$}.
\end{align}
Then, for a sufficiently small $p>0$, we have $\BIR_2(n)>\BIR_1(n)$ for all local steps $n\geq n_0$.
\end{example}

\begin{proof}
By definition of the modified algorithm we have that
\begin{align}\label{eq:random-BIR}
\BIR_2(n)
    &= \Ex{m\sim (n_0-1)+\term{Binomial}(n-n_0+1,1-p)}{(1-p) \cdot \BIR_1(m) + p \cdot \BIRgr(m)}.
\end{align}
In this expression, $m$ is the number of times \alg[1] is invoked in the first $n$ steps of the modified algorithm. Note that
    $\E[m] = n_0-1 + (n-n_0+1)(1-p) \geq (1-p)n$.

Then for all $n\geq n_0$ and small enough $p>0$ it holds that:
\begin{align*}
 \BIR_2(n)
    &\geq  (1-p+pc)\; \E[\; \BIR_1(m) \;] \\
\E[\; \BIR_1(m) \;]
    &\geq \BIR_1(\; \E[m] \;) &\qquad\text{(By Jensen's inequality)} \\
    &\geq \BIR_1(\; (1-p)n \;) &\qquad\text{(since $\E[m]\geq n(1-p)$)}  \\
    &\geq \beta\cdot n^{-\gamma} \cdot (1-p)^{-\gamma}
        &\qquad\text{(plugging in $\BIR_1(n)=\beta n^{-\gamma}$)}  \\
    &> \BIR_1(n)\;\; (1-p\gamma)^{-1}
        &\qquad\text{(since $(1-p)^\gamma < 1-p\gamma$)}.\\
\BIR_2(n)
    &>\alpha\cdot \BIR_1(n),
    &\text{where} \quad
    \alpha = \frac{1-p(1-c)}{1-p\gamma}>1.
\end{align*}
(In the above equations, all expectations are over $m$ distributed as in \eqref{eq:random-BIR}.)
\end{proof}

@@

\begin{theorem}\label{thm:SoftMax-strong}
Assume the \SoftMaxRandom response function. Suppose algorithms \alg[1], \alg[2] are monotone and $T$-oblivious, and \alg[1] weakly-BIR-dominates \alg[2]. Posit mild technical assumptions:
  $\BIR_1(n) \to 0$ and
\begin{align}\label{eq:SoftMax-assn-strong}
 % (\forall n\geq n(\eps)) \qquad
 %  \BIR_2(n) > e^{-\eps n}. %used to be 4exp( -\eps n/6)
(\exists n_0\; \forall n\geq n_0) \qquad
\BIR_2(n) \geq \frac{4}{\alpha_0}
\exp \left( \frac{-n\,\min\{\eps_0, 1/8\}}{12}\right).
\end{align}
Then there
  exists some $t_0$ such that each agent $t\geq t_0$ chooses principal
  $1$ with probability
\begin{align}\label{eq:thm:SoftMax-strong}
     \Pr[i_t = 1]\geq \tfrac12 +  \tfrac14 \,c_0\,\alpha_0\; \BIR_2(t).
\end{align}
\end{theorem}

@@

\begin{corollary}\label{cor:SoftMax-strong}
Assume \SoftMaxRandom response function. Consider the \FiniteGame in which all feasible algorithms satisfy $\BReg(n)\to \infty$.%
\footnote{$\BReg(n)\to \infty$ is a mild non-degeneracy condition.
\ascomment{TODO: can we actually prove it? Assume Bernoulli priors, discrete prior, reduce to Lai-Robbins.}}
Suppose a feasible algorithm \alg BIR-dominates all others. Then, for any sufficiently large time horizon $T$, there is a unique Nash equilibrium: both principals choose \alg.
\end{corollary}

>>>>

We characterize the inherent quality of an MAB algorithm in terms of its \emph{Bayesian Instantaneous Regret} (henceforth, \BIR), a standard notion from machine learning:
\begin{align}\label{eq:theory-BIR-defn}
\BIR_i(n) := \E\sbr{ n\cdot \max_{a\in A} \mu_a - \rew_i(n)}.
\end{align}
We are primarily interested in how \BIR scales with $n$: typically, how fast \BIR decreases with $n$. We treat $K$, the number of arms, as a constant unless specified otherwise. Intuitively, (much) better MAB algorithms tend to have a (much) smaller BIR, see Appendix~\ref{app:examples} for more background.

%Let $\BIR_i(n) = \BIR(n)$ for the algorithm of principal $i$.

@@

@@

%\ascomment{Steven: OLD EXAMPLE. I revised it to clarify what it is supposed to mean. But my new example is better, isn't it?}
%
%\ascomment{The conclusion can't be right: "greedy modification" can only improve the BIR at step $n_0$! Perhaps what we derive is that
%$\BIR_2(n)>\BIR_1(n)$ starting from $n=n_0+1$? }
%
%
%\begin{example}
%Consider \alg[1] and \alg[2] as in Theorem~\ref{thm:random-greedy}.
%Let $\BIRgr(n)$ be the \BIR of the ``greedy choice" after $n-1$ steps of \alg[1]. Assume, for some absolute constants $\beta,\gamma>0$ and $c>1-\gamma$, that
%\begin{align}\label{eq:cor:random-greedy-ex-assumption}
% \tfrac{1}{c}\; \BIRgr(n) \geq \BIR_1(n)= \beta\cdot n^{-\gamma}
% \qquad \text{for all local steps $n$}.
%\end{align}
%Then, for a sufficiently small $p>0$, we have $\BIR_2(n)>\BIR_1(n)$ for all local steps $n\geq n_0$.
%\end{example}
%
%\begin{proof}
%By definition of the modified algorithm we have that
%\begin{align}\label{eq:random-BIR}
%\BIR_2(n)
%    &= \Ex{m\sim (n_0-1)+\term{Binomial}(n-n_0+1,1-p)}{(1-p) \cdot \BIR_1(m) + p \cdot \BIRgr(m)}.
%\end{align}
%In this expression, $m$ is the number of times \alg[1] is invoked in the first $n$ steps of the modified algorithm. Note that
%    $\E[m] = n_0-1 + (n-n_0+1)(1-p) \geq (1-p)n$.
%
%
%Then for all $n\geq n_0$ and small enough $p>0$ it holds that:
%\begin{align*}
% \BIR_2(n)
%    &\geq  (1-p+pc)\; \E[\; \BIR_1(m) \;] \\
%\E[\; \BIR_1(m) \;]
%    &\geq \BIR_1(\; \E[m] \;) &\qquad\text{(By Jensen's inequality)} \\
%    &\geq \BIR_1(\; (1-p)n \;) &\qquad\text{(since $\E[m]\geq n(1-p)$)}  \\
%    &\geq \beta\cdot n^{-\gamma} \cdot (1-p)^{-\gamma}
%        &\qquad\text{(plugging in $\BIR_1(n)=\beta n^{-\gamma}$)}  \\
%    &> \BIR_1(n)\;\; (1-p\gamma)^{-1}
%        &\qquad\text{(since $(1-p)^\gamma < 1-p\gamma$)}.\\
%\BIR_2(n)
%    &>\alpha\cdot \BIR_1(n),
%    &\text{where} \quad
%    \alpha = \frac{1-p(1-c)}{1-p\gamma}>1.
%\end{align*}
%(In the above equations, all expectations are over $m$ distributed as in \eqref{eq:random-BIR}.)
%\end{proof}

@@

\begin{theorem}\label{thm:SoftMax-strong}
Assume the \SoftMaxRandom response function. Suppose algorithms \alg[1], \alg[2] are monotone and $T$-oblivious, and \alg[1] weakly-BIR-dominates \alg[2]. Posit mild technical assumptions:
  $\BIR_1(n) \to 0$ and that $\BIR_2$ cannot be extremely small, namely: 
\begin{align}\label{eq:SoftMax-assn-strong}
 % (\forall n\geq n(\eps)) \qquad
 %  \BIR_2(n) > e^{-\eps n}. %used to be 4exp( -\eps n/6)
(\exists m_0\; \forall n\geq m_0) \qquad
\BIR_2(n) \geq \frac{4}{\alpha_0}
\exp \left( \frac{-n\,\min\{\eps_0, \nicefrac{1}{8}\}}{12}\right).
\end{align}
Then there
  exists some $t_0$ such that each agent $t\geq t_0$ chooses principal
  $1$ with probability
\begin{align}\label{eq:thm:SoftMax-strong}
     \Pr[i_t = 1]\geq \tfrac12 +  \tfrac14 \,c_0\,\alpha_0\; \BIR_2(t).
\end{align}
\end{theorem}

@@

\begin{corollary}\label{cor:SoftMax-strong}
Assume \SoftMaxRandom response function. Consider the \FiniteGame in which all feasible algorithms satisfy $\BReg(n)\to \infty$.%
\footnote{$\BReg(n)\to \infty$ is a mild non-degeneracy condition, see 
\ascomment{insert pointer}.}
Suppose a feasible algorithm \alg BIR-dominates all others. Then, for any sufficiently large time horizon $T$, there is a unique Nash equilibrium: both principals choose \alg.
\end{corollary}
